---
title: "Assignment 10Dply B: Comparison between dplyr and SQL SELECT"
author: "Joshua Conte"
date: "July 16, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Comparison between dplyr and SQL SELECT
For this assignment, I need to: extract aggregated statistics using dplyr and compare dplyr to SQL SELECT.

##Setting up R
First I configure R studio with the parameters below:
```{r Configure RStudio}
# clears the console in RStudio
cat("\014") 

# clears environment
rm(list = ls())  

# Set working directory
setwd("C:/R/DA5020/Week_10/")

# Load required packages
library(dplyr)
```

#Begin Assignment 10Dply B
##Assignment Tasks
###Task 1
Revisit your Bird Strike database created for the last assignment. Connect to the database.
```{r Task 1}
# connect to the database file using src_sqlite
birdstrike<- src_sqlite("birdstrike.db", create = F)

# List the tables in the database
src_tbls(birdstrike)
```

**Querying the database** \
dplyr translates the R code we write to SQL code. Inorder to connect to the tables, I needed to use 'tbl', as shown below.

For table1:
```{r table1}
# Load table1
table1 <- tbl(birdstrike,"table1")

# This shows the first 3 rows of table1
head(table1,3)
```

For table2:
```{r table2}
# Load table2
table2 <- tbl(birdstrike,"table2")

# This shows the first 3 rows of table2
head(table2,3)
```

###Task 2
Write a dpyr statement that counts the number of incidents where the incident reported fog during the incident.

I can pipe dplyr operations together with %>%. The pipeline %>% takes the output from the left-hand side of the pipe as the first argument to the function on the right hand side.
```{r Task 2}
# Grouped by ConditionsPrecipitation and found the count for each 
# ConditionsPrecipitation type as BirdStrikes.  Then filtered for Fog.
table2%>%group_by(ConditionsPrecipitation)%>% 
  summarize(BirdStrikes=n())%>%      
  filter(ConditionsPrecipitation=='Fog')
```

###Task 3
Write a function **CountIncidents(aircraftType)** that accepts an aircraft type and
returns the number of bird strike incidents for the aircraft type.
```{r Task 3}
CountIncidents <- function(x) {
  # This pipes dplyr operations together with %>%. 
  # Grouped by AircraftType, then found the count for each AircraftType type as 
  # BirdStrikes. Then filtered for 'x'.
  table1 %>% group_by(AircraftType) %>%
    summarize(BirdStrikes = n()) %>%
    filter(AircraftType == x)
}

# This calls the function with argument in quotes
CountIncidents("Helicopter")
```

###Task 4
Write a function **Incidents(Airline)** that accepts an airline and returns a dataframe that contains all incidents for that airline. Limit the columns to: **AirportName**, **AircraftModel**, and **FlightDate**.
```{r Task 4}
Incidents <- function(x) {
  # THis function accepts an airline and returns a dataframe that contains all
  # incidents for that airline. Limiting the columns to: AirportName, AircraftModel,
  # and FlightDate
  
  # This pipes dplyr operations together with %>%.
  # Select AircraftAirlineOperator, AirportName, AircraftMakeModel, FlightDate then
  # filtered for AircraftAirlineOperator 'x'. Ungrouped AircraftAirlineOperator,
  # then arrange by date oldest to newest
  table1 %>% select(AircraftAirlineOperator,
                    AirportName,
                    AircraftMakeModel,
                    FlightDate) %>%
    filter(AircraftAirlineOperator == x) %>%
    ungroup() %>%
    select(-AircraftAirlineOperator) %>%
    arrange(FlightDate)
}

# This calls the function with argument in quotes
Incidents.df <- Incidents("CONTINENTAL AIRLINES")

# Make it a data frame
Incidents.df <- data.frame(Incidents.df)

# Summarize the data frame
summary(Incidents.df)

# Show first three lines
head(Incidents.df, 3)
```

###Task 5
Write a function **CountIncidentsByAirline()** that creates a data frame where the first column is a name of an Aircraft and the second column is the total number of incidents the Airline had.
```{r Task 5}
CountIncidentsByAirline <- function(x) {
  # THis function accepts an airline and returns a dataframe that contains all 
  # incidents for that airline. Limiting the columns to: AirportName, AircraftModel, 
  # and FlightDate
  
  # This pipes dplyr operations together with %>%. 
  # Select AircraftAirlineOperator, AircraftMakeModel, then filtered for 
  # AircraftAirlineOperator 'x', ungrouped AircraftAirlineOperator.
  # Then group by AircraftMakeModel and summarize by birdstikes and
  # order data by largest to smallest birdstrikes.
  table1 %>% select(AircraftMakeModel) %>%
  group_by(AircraftMakeModel) %>%
  summarize(BirdStrikes = n()) %>%
  arrange(desc(BirdStrikes)) 
}

# This calls the function with argument in quotes
IncidentsByAirline.df <- CountIncidentsByAirline()

# Make it a data frame 
IncidentsByAirline.df <- data.frame(IncidentsByAirline.df)

# Summarize the data frame
summary(IncidentsByAirline.df)

# Show first three lines
head(IncidentsByAirline.df,3)
```

###Task 6:
Use **show_query()** to extract the SQL statement from one of your dplyr operations, then compare it with the corresponding SQL statement you wrote last week. State your preference for retrieving data from a large data set.\

I selected the statement from task2, a statement that counts the number of incidents where the incident reported fog during the incident:
```{r Task 6}

fog_dplyr<-table2%>%group_by(ConditionsPrecipitation)%>% 
  summarize(BirdStrikes=n())%>%      
  filter(ConditionsPrecipitation=='Fog')

show_query(fog_dplyr)
```
**SQL statement wrote last week:** \
SELECT ConditionsPrecipitation, count(*) AS BirdStrikes \
FROM table2 \
WHERE ConditionsPrecipitation \
LIKE 'fog' \


The statements are similar, however, the one I used last week was a little smaller since I utilized the LIKE statement.  Overall, I prefer to use SQL over dplyr because there is so much more support for SQL.  It took me a while to find the basics of dplyr, but for SQL, it's just a quick search on stackoverflow.  Plus, SQL is used in a lot of different places, so it's easier to port the code from one place to another.